% 1M skyldes lønn til mennesker eller kun compute? (tid = penger hvis det er
% mennesker)

% BF: Har du en referanse vi kan bruke her? Dette er en ganske vanlig quote, men
% husker ikke hvor jeg har sett den, Dette er veldig generelle greier, men vi
% skulle hatt et par gode paper å sitere. 
% LAB: Rob Finn (EBI) bruker denne: https://genomebiology.biomedcentral.com/articles/10.1186/gb-2011-12-8-125
% LAB: Også har vi disse: https://www.veritasgenetics.com/mygenome
In the past decade the generation of biological datasets has been unprecedented,
and the famous "\$1000k genome, \$1M analysis"\cite{} has become more apparent.
% mer om biologi og data 

% Trenger litt tekst som sier at vi skal snakke om stats, presentasjon, dbs,
% microservices og kobling av alle. 

% R og stats 
% LAB: To analyze eller To explore
To analyze the growing biological data sets, there is now a number of analysis
tools in various programming languages.\cite{} In R, there are popular package
repositories such as CRAN \url{cran.r-project.org} or Bioconductor
\url{bioconductor.org} where developers can share software packages and keep
them up-to-date.  These repositories contain software packages for exploring
high-throughput genomic data in one environment. This includes both
pre-processing, e.g. cleaning, removing outliers, and analysing it with known
statistical methods. In other languages such as Python or Go developers can
choose bioinformatics libraries such as BioPython\cite{} and biogo\cite{}
respectively.  To use these packages to their full potential researchers require
a high level of coding skill, and 
% LAB: Omskrive: To fully utilize these packages requires ...coding skills LAB:
% ...og selv da er det ikke alt som enkelt kan løses, som sklaering i R, etc
% LAB: Og hva er problemet og løsningen for dette (vis tools?) 
However all of these software packages require users to have a
high level of coding skills. 

% Vis 
% LAB: Henger ikke sammen med forrige
% LAB: Bedre å starter med denne? Vis tools er løsningen, hvis de finnes for
% problemet. Hvis ikke må sette i gang og kode.
A key part of analysing biological data is to visualize it. Researchers often
start data analysis by using simple visualization techniques to get a quick
overview of the data. Continuing in the data analysis pipeline researchers can
use more advanced visual techniques to explore the data either using
software packages or complete data visualization applications. Traditionally
data visualization applications have been built as desktop applications that
require installation and setup by the user, but now the move is towards software
that run in the web browser without any user setup. 

% WIP WIP WIP
% In systems biology there is a need for visualization tools that can make use
% of the lastest statistical packages and interact with these. It is not enough
% to visualize the results of an analysis pipeline, but interact with the
% different steps from the final visualization. Examples of this interaction can
% be modifying clustering parameters that are used in the analysis.

% Data fra R til vis
To visualize and share results from statistical analyses, the results are often
exported to a data format such as comma-separated values (CSV) and
then visualized using an external tool. This decouples data presentation and
analysis.  
% LAB: Hva er bra/dårlig med decoupling
Through initiatives such as ApacheR and OpenCPU there has been a move
towards embedding scientific computation in data visualization application. This
removes the decoupling of statistical frameworks and interaction with the
analysis results. 

% Koble data sammen med db. 
% LAB: Hva er sammenhengen med programming packages & viz tools?
Interpreting the analysis results require integration of known biology, either
from biological databases such as MSigDB\cite{} or KEGG\cite{}, or through
scientific publications from reference databases such as PubMed\cite{}.
Performing manual lookup into these databases is often tedious and error-prone,
making it necessary to automate the task. Now as more databases provide REST
APIs it is possible to provide software packages for automatic retrival of
database information. In addition, since databases are continuously being
updated, using a REST API to retrieve database information will ensure that the
data is always kept up to date. 

% Hvorfor bryr vi oss om både stats + DBs?  Skal vi si noe om at det blir
% vanligere å bruke REST APIs fremfor å laste ned alt og lese ting lokalt

% Lage apps som gjør alt dette. 

% Hva er micro-services? 
% LAB: Hadde forventet en beskrivelse av ett problem (motivert i paragrafene
% overnfor), som microservices løser
Microservice architectures structures applications into small reusable, loosely
coupled parts. These communicate via lightweight programming language-agnostic
protocols such as HTTP, making it possible to write single applications in
multiple different programming languages. This makes it possible to use the most
suitable programming language for each specific part. E.g. to use R and
Bioconductor to analyse biological data, or C++ and OpenCV for high-performance
computer vision tasks, or HTML, CSS, and JavaScript to build portable
user-interfaces. To build a microservice application, developers bundle each
service in a container that are deployed. Containers are built from
configuration files which describe the operating system, software packages and
versions of these. This makes reproducing an application a trivial task. The
most popular implementation of a software container is Docker\footnote{\url{}},
but others such as Rkt\footnote{\url{}} exist.  

% LAB: Related bio work? Annen microservice related work? Hvorfor ikke bare bruke de?


\subsubsection*{Requirements} 
From our experience building data exploration applications we have identified

... the
following requirements 

\begin{enumerate}
    \item A language-independent approach for integrating, or embedding,
        statistical software, such as R, directly in interactive data
        exploration applications.
    % Ikke motivert behovet for up-to-date
    \item A 
        
        Up-to-date interfaces to online databases providing meta-data for
        understanding results from statistical analyses.
    % Ikke sagt noe om maintance & deployment
    \item Ease of software development, maintenance and deployment. 
\end{enumerate} 

\subsubsection*{Contributions} 
Our contributions are: 
\begin{enumerate}
% men ikke visualization?
\item An approach for developing data exploration applications for systems
biology that combine statistical analyses with online databases.  
\item A demonstration of its viability through N different applications. 
\item Performance evaluation of its central data analysis component. 
\end{enumerate} 


\section*{Related Work} 
% Hva er korrekt: "in" system biology eller "for" system biology?
In this section we aim to cover the existing approaches for building interactive
data exploration applications in systems biology. 

% TODO: Add related work where somebody has built a system for exploring some
% data where 

% Det er en utfordring å ha related work tidlig i dette paperet. Slik det er nå,
% er det ikke sikkert leseren skjønner hvorfor diss systemene diskuteres, hva
% som er sammenhengen mellom de, og hva det har å gjrøe med Kvik.

\subsubsection*{OpenCPU} 
OpenCPU is a system for embedded scientific computing and reproducible
research.\cite{opencpu} It provides an HTTP API to the R programming language to
provide an interface with statistical methods. It enables users to make function
calls to any R package and retrieve the results in a wide variety of formats
such as json or pdf. 
Users can chose to host their own R server or use public servers, and OpenCPU
works in a single-user setting within an R session, or a multi-user setting
facilitating multiple parallel requests. 
OpenCPU provides a Javascript library for interfacing with R, as well as Docker
containers for easy installation. OpenCPU has been used to build multiple
applications.\footnote{\url{opencpu.org/apps.html}}.
% Design patterns er ikke nevnt enda
In Kvik we provide a package
to interface with OpenCPU servers from the Go programming language since it
follows the design pattern we have chosen to interface with data and analyses. 


\subsubsection*{Renjin} 
% Skjønner ikke poenget med Renjin. Dvs hva fordelen med å kjøre R kode på en
% JVM er.
Renjin is a JVM-based interpreter for the R programming language.\cite{renjin}
It targets developers who want to integrate the R interpreter in web
applications. Since it is built on top of the JVM it allows developers to write
data exploration applications that interact directly with R code. Although
Renjin supports a large number of CRAN packages it cannot access any R package
(e.g. from BioConductor) without modification. This makes the programming effort
to use Renjin as an interface with R higher. 

\subsubsection*{Shiny} 
Shiny is a web application framework for R.\cite{shiny} It allows developers to
build web applications in R without having to have any knowledge about HTML, CSS
or Javascript. It provides a widget library to provide more advanced Javascript
visualizations such as Leaflet for maps or threejs for WebGL-accellerated
graphics. Developers can choose to host their own web server with the user-built
Shiny Apps, or host them on public servers. Shiny forces users to implement data
exploration applications in R, limiting the functionality to the 
widgets and libraries in Shiny. 

% Synes SparkR burde nevnes

\subsubsection*{Biogo} 
Biogo is a bioinformatics library in Go. It provides functionality to analyse
genomic and metagenomc datasets in the go programming
language.\cite{Kortschak005033} Using the go programming language the developers
are able to provide high-performance parallel processing in a clean and simple
programming language. 

\subsubsection*{Cytoscape} 
Cytoscape is an open source software platform for visualizing complex
networks and integrating these with any type of attribute
data\cite{shannon2003cytoscape}. It allows for analysis and visualization in the
same platform. Users can add additional features, such as databases connections
or new layouts, through Apps. To bring the visualization and analysis
capabilities to the web the creators of Cytoscape have developed
Cytoscape.js\footnote{\url{js.cytoscapejs.org}}, a Javascript library to create
interactive graph visualizations. 

cyREST. 



