In recent years the biological community has generated and collected an
unprecedented ammount of data. While the cost of data collection has
drastically decreased, data analysis continue to be a larger fraction of the
total experiment cost.\cite{sboner2011real}  An important part of data analysis
includes the time spent by human experts interpreting the results. This calls
for novel methods in data analysis and exploration. 

Several tools are now available in various programming languages for biological
data analysis. These include novel bioinformatics methodologies and graphical
analysis tools.
In the R statistical programming language developers share software through
repositories such as CRAN\footnote{\url{cran.r-project.org}} or
Bioconductor\footnote{\url{bioconductor.org}}.  In other languages, libraries
for biological bomputation are often availalbe like BioPython\cite{biopython}
and biogo\cite{biogo} for Python or Go, respectively. Projects such as
Galaxy\footnote{\url{galaxyproject.org}} and Common Workflow Language
(CWL)\footnote{\url{commonwl.org}} enable resarchers to build and run
biological data analysis pipelines consisting of a wide range of tools.
Although these framework are tremendously helpful, we need novel approaches
to integrate them with interactive visualizations and clean user-interfaces to
make them more accessible to the researchers interpreting the results. 

In the field of systems biology, data exploration applications need to link
results to relevant prior knowledge. In the later years there has been a
tremendous effort to curate databases with relevant information on genes and
processes. Databases such as the Molecular Signatures Database
(MSigDB)\footnote{\url{software.broadinstitute.org/gsea/msigdb}} and the Kyoto
Encyclopedia of Genes and Genomes (KEGG)\footnote{\url{kegg.jp}} both provide
interfaces to retrieve data that can be used to better the understanding of
data analysis results. 

While it would make development easy, we argue that data exploration
applications in systems biology should not be written in a single programming
language.  Different programming languages solve different tasks.  For example,
new biological data analysis techniques are quickly realeased in R and its
package repositories; high-performance computer vision tasks are performed in
C++ and OpenCV; and portable user interfaces more easily built in HTML, CSS and
JavaScript.  Therefore applications that integrate novel statistical analysis
tools, interactive visualizations, and biological databases likely need to
include several components written in different languages. 

% LAB: dette blir vel "muntlig". Jeg synes at dette er stedet å løfte opp
% utfordringene med å utvikle ordentlig software, dvs:
% 1. Noe som virker
% 2. Som har god nok performance
% 3. Hvor det er mulig etterpå å forstå hva analysen gjorde
% 4. Som skalerer til større datasett
% 5. Som er mulig å vedlikeholde etterpå
% 6. Som har deler som kan gjennbrukes av andre
% Ofte er det bare 1. som blir gjort. Men feks for oss i Elixir som skal tilby
% tjenester som faktisk virker, så er det de andre punktene som er viktig.

A microservice architecture structures an application into small reusable,
loosely coupled parts. These communicate via lightweight programming
language-agnostic protocols such as HTTP, making it possible to write single
applications in multiple different programming languages. This way the
most suitable programming language is used for each specific part. To build a
microservice application, developers bundle each service in a container.
Containers are built from configuration files which describe the operating
system, software packages and versions of these. 
This makes reproducing the analyses, database lookups, library versions in an
application a trivial task. The most popular implementation of a software
container is Docker\footnote{\url{docker.com}}, but others such as
Rkt\footnote{\url{coreos.com/rkt}} exist. Initiatives such as
BioContainers\footnote{\url{biocontainers.pro}} now provide containers
pre-installed with different bioinformatics tools. 

To build these services we need a framework that fulfills the following
requirements: 

\begin{enumerate}
    \item It provides a language-independent approach for integrating, or
        embedding, statistical software, such as R, directly in 
        interactive data exploration applications.
    \item It has a low latency language-independent interface to online
        reference databases in biology that captures provenance of the
        meta-data. 
    \item It provides software packages for building microservices that are easy
        to develop and maintain. 
    \item Its components are easy to deploy and share between applications and
        projects. 
    \item Its components provide low response times necessary to develop
        interactive applications, and makes it easy to scale out to many users
\end{enumerate} 

First, we identified a set of reusable services that application developers can
use to build a wide range of applications. The key services of a biological data
exploration application are i) a compute service for executing statistical
analyses in languages such as R, and ii) a database query service for retrieving
information from biological databases. We believe that the user-facing logic is
best left out to the application developer, providing language-indenpendent
interfaces to both services. 

In this paper, we describe a novel approach for building data exploration
applications in systems biology. We show that by building applications as a set
of services we can reuse and share its components between
applications. In addition, by packaging the services using container technology
we promote reproducible research and simplify application deployment. We have
used our approach to build a number of applications, both command-line and
web-based. In this paper we describe how we used our approach to develop MIxT,
a web application for exploring and comparing transcriptional profiles from
blood and tumor samples. The MIxT web application integrates statistical
analysis together with biological databases and interactive visualizations.

% LAB: Related bio work? Annen microservice related work? Hvorfor ikke bare
% bruke de?
