% TODO: 
% - Draw up a sketch of what is MIxT and what is Kvik
% - What interfaces does Kvik provide? 
% - Spark, look at their papers and the app / abstractions. 
% - What are the ops done most frequent? Get data from KEGG/Msigdb etc? 
% - Scrap hard line breaks for LA (soft lines instead) 
% - add pdf to git. 




\section*{Background}
% Snakk litt om de forskjellige systemene i hver disciplin
% --> Standing on the shoulders of giants 
% Analysis frameworks: R, Bioconductor, Galaxy
% Visualization: Cytoscape, BioJS, D3
% Databases: MSiGDB, KEGG, GeneCards, PubMed
% Parallelized and distributed computation: Hadoop, Spark, Kubernets, Azure
Analysis frameworks: Bioconductor in R, Python.
Visualization: Cytoscpape, D3, BioJS.
Databases: MSigDB, KEGG, PubMed.
Parallel and distributed: Hadoop, Spark, Kubernetes, Azure. 

Data wrangling and analysis done in R or other languages.
Visualization and presentation of final datasets by external tools.
Manual database lookup is tedious and gets out of date.

Related work: OpenCPU, biogo, renjin, 

\emph{The big issue: There is no link between analysis and final presentation} 

\section*{Methods}
Collection of packages in the Go programming language for building data
exploration applications.  Interfaces with popular online databases such as
MSigDB and KEGG.  Provides an interface to the R programming language.
Typically used to build web apps, but commandline tools are also possible!


\section*{Results and Discussion}
MIxT. 
Interactive web application for exploring WGCNA modules from blood and biopsy.
Integrate results with known genesets and online databases.
Docker containers for reproducibility and portability.

Data analysis: R package  + Kvik = R in the browser. 
Visualization: D3, SigmaJS, R Plots 

How can we improve Kvik and its features: R user input sanitation, security, 


\section*{Conclusions}
Sum it all up. 
